## QIIME2 Workflow

QIIME2 is a powerful, scalable and decentralised microbiome analysis platform that brings together a wide range of analytical capabilities, and has been online since 2019.

More information in :

-   Please cite: reproducible, interactive, scalable and extensible microbiome data science using QIIME 2. Nature Biotechnology. 2019, 37(8): 852-857. <https://docs.qiime2.org/2023.5>\
-   Check out the QIIME2 website for details on how to install it! <https://docs.qiime2.org/2023.2/install/>
-   Check also <https://xkcococo.github.io/blog/2023/Use-QIIME2-to-Preprocess-16S-Sequencing-Files/> a very good site for importing 16S

## Load the QIIME2

```{r, InstalationQIIME2, eval=F,echo=T}
export PATH=~/miniconda3/bin:$PATH #select path
source activate qiime2-2023.5

# MoreTmpSpace
export TMPDIR=/home/PJJ/tmp 
```

## Importing Data into QIIME2

::: callout-warning
Underscore incompatible with import into qiime2
:::

for delet underscore
```{r , eval=F}
for file in *_*; do mv "$file" "${file//_/}"; done
```


## Create manifest for QIIME2

```{r, CreationManifest, eval=F,echo=T}
cd /home/cmaslard/data_qiime2/seq
echo 'sample-id','absolute-filepath','direction'> manifest.txt
ls *.fastq|while read id; #add .gz if compress
do
echo "${id%%_*},$PWD/$id,forward">> /home/cmaslard/data_qiime2/seq/manifest.txt;
done
```

## Import sequence in QIIME2 (demux)

**This step merges all the samples together, the output file is the merged paired-demux.qza for denoising.** No need to trim the adapters because i use **clean data**.

```{r ,ImportationToQIIME2 ,eval=F,echo=T}
cd /home/PJJ/CM/data
qiime tools import \
--type 'SampleData[SequencesWithQuality]' \
--input-path manifest.txt \
--output-path demux.qza \
--input-format SingleEndFastqManifestPhred33 
```

## Sequence Denoising with DADA2

Dada2 denoising generates ASVs and representative sequences, the input file paired-demux.qza used in the previous step, and the output file is the ASV table table-dada2.qza in qza format and representative sequences rep-seqs-dada2.qza

```{r ,DenoisingWithDADA2 ,eval=F,echo=T}
time qiime dada2 denoise-single \
--i-demultiplexed-seqs demux.qza \
--p-trim-left 0 \
--p-trunc-len 120 \
--o-representative-sequences rep-seqs-dada2.qza \
--o-table table-dada2.qza \
--o-denoising-stats stats-dada2.qza 
```

## Generating the Taxonomy Table

There are two options for generating the taxonomy table: 1. Train your own feature classifier, or 2. Use a pre-traiend feature classifier provided by QIIME.

### For Bacteria

I will use the pre-trained classifier "Silva 138 99% OTUs full-length sequences.

For more information see :

-   <https://www.arb-silva.de/documentation/release-138/> #silva-138
-   <https://zenodo.org/record/6395539> #silva-138-99-nb-classifier.qza
-   <https://docs.qiime2.org/2023.9/data-resources/#taxonomy-classifiers-for-use-with-q2-feature-classifier>

```{r ,download classifier ,eval=F,echo=T}
#wget -O "silva-138-99-nb-classifier.qza" "https://data.qiime2.org/2020.8/common/silva-138-99-nb-classifier.qza" #old version
wget -O "silva-138-99-nb-classifier.qza" "https://data.qiime2.org/2023.9/common/silva-138-99-nb-classifier.qza"
```

```{r, ProcessClassifier16S, eval=F,echo=T}
qiime feature-classifier classify-sklearn \
	  --i-classifier silva-138-99-nb-classifier.qza \
	  --i-reads rep-seqs-dada2.qza \
	  --o-classification taxonomy.qza
```

### For Fungi

I will use the classifier "Unit V9 99%. find in [UNITE QIIME release for Fungi 2 V9](https://dx.doi.org/10.15156/BIO/2938080) publish in @abarenkov_unite_2024

For tutoral see [Training the QIIME2 Classifier with UNITE ITS Reference Sequences](https://john-quensen.com/tutorials/training-the-qiime2-classifier-with-unite-its-reference-sequences/) for train classifier. *Instructions for creating a classifier file to be used by QIIME2 for the classification of fungal ITS sequences.*

You can use also the pretrained classifier from [git hub UNITE v9.0 v25.07.2023 for qiime2-2023.9](https://github.com/colinbrislawn/unite-train/releases/tag/v9.0-v25.07.2023-qiime2-2023.9)

-   Import the UNITE reference sequences into QIIME2.

```{r, ImportClassifier, eval=F,echo=T}
qiime tools import \
--type FeatureData[Sequence] \
--input-path sh_refs_qiime_ver9_99_s_25.07.2023.fasta \
--output-path unite-ver9-seqs_99_25.07.2023.qza
```

-   Import the taxonomy file.

    ```{r, ImportTaxoITS, eval=F,echo=T}
    qiime tools import \
    --type FeatureData[Taxonomy] \
    --input-path sh_taxonomy_qiime_ver9_99_s_25.07.2023.txt \
    --output-path unite-ver9-taxonomy_99_25.07.2023.qza \
    --input-format HeaderlessTSVTaxonomyFormat
    ```

-   Train the classifier

    ```{r, TrainClassifier, eval=F, echo=T}
    qiime feature-classifier fit-classifier-naive-bayes \
    --i-reference-reads unite-ver9-seqs_99_25.07.2023.qza \
    --i-reference-taxonomy unite-ver9-taxonomy_99_25.07.2023.qza \
    --o-classifier unite-ver9-99-classifier-25.07.2023.qza
    ```

-   Process classifier for ITS

    ```{r, ProcessClassifierITS, eval=F,echo=T}
    qiime feature-classifier classify-sklearn \
    	  --i-classifier unite_ver9_99_s_25.07.2023-Q2-2023.9.qza \
    	  --i-reads rep-seqs-dada2.qza \
    	  --o-classification taxonomy.qza
    ```

## Generate a phylogenetic tree

Generate a phylogenetic tree is important for calcul UniFrac diversity

**Generate a Multiple Sequence Alignment (MSA)** Use qiime alignment mafft or another aligner to align your representative sequences. This step is crucial for constructing a phylogenetic tree.

```{r, GenerateMSA, eval=F,echo=T}
qiime alignment mafft \
  --i-sequences rep-seqs-dada2.qza \
  --o-alignment aligned-rep-seqs.qza
```

**Mask the Alignment** Masking the alignment helps to remove poorly aligned regions, which can negatively impact tree inference.

```{r, MaskAlignment, echo=T, eval=F}
qiime alignment mask \
  --i-alignment aligned-rep-seqs.qza \
  --o-masked-alignment masked-aligned-rep-seqs.qza
```

**Construct a Phylogenetic Tree** Use a phylogenetic tree-building method, such as FastTree or RAxML.

```{r, MakePhylogeneticTree,echo=T, eval=F }
qiime phylogeny fasttree \
  --i-alignment masked-aligned-rep-seqs.qza \
  --o-tree unrooted-tree.qza
```

Optionally, you can root the tree using midpoint rooting.

```{r, RootedTree,echo=T, eval=F}
qiime phylogeny midpoint-root \
  --i-tree unrooted-tree.qza \
  --o-rooted-tree rooted-tree.qza
```

```{r, ExportTaxonomy, eval=F,echo=T}
#Export taxonomy info in .tsv format
qiime tools export \
  --input-path taxonomy.qza \
  --output-path exported-feature-table

#Output visualizations:
qiime metadata tabulate \
  --m-input-file taxonomy.qza \
  --o-visualization taxonomy.qzv
```

## Export results

In this step, we will use table-dada2.qza to generate our feature table. Export feature table to biom format.

```{r, ExportFeatureTable, eval=F,echo=T}
#Creating a TSV BIOM table
#first, export your data as a .biom
qiime tools export --input-path table-dada2.qza --output-path exported-feature-table

#Convert .biom to .tsv
biom convert -i exported-feature-table/feature-table.biom -o exported-feature-table/feature-table.tsv --to-tsv
biom head -i feature-table.tsv

biom convert -i feature-table/feature-table.biom \
      -o feature-table/feature-table.txt \
      --to-tsv

# Delete comment lines
  ed -i '/# Const/d' feature-table/feature-table.txt

# For me, i prefer import in R via .csv
biom convert -i exported-feature-table/feature-table.biom -o exported-feature-table/feature-table.csv --to-tsv

# Exporting species annotations
qiime tools export \
      --input-path taxonomy.qza \
      --output-path taxonomy

# measure unifrac distance
qiime diversity-lib weighted-unifrac \
    --i-table table-dada2.qza \
    --i-phylogeny rooted-tree.qza \
    --p-threads 100 \
    --o-distance-matrix weighted-unifrac-dm.qza

# Export the distance matrix from QIIME 2:
qiime tools export --input-path weighted-unifrac-dm.qza --output-path unifracDist

# Convert the exported distance matrix to a csv file:
biom convert -i unifracDist/distance-matrix.tsv --to-tsv --header-key taxonomy -o unifracDist.csv
```
